Elderly, Male, Sporadic Population is More Likely to Get Severe COVID-19

### 2022-04-27 outputs/paper/paper.Rmd
final version

---
title: "Estimation of COVID-19 cases in Ontario"
author:
- Zihan Zhang
thanks: "Code and data are available at: https://github.com/zihanjasmine/Estimation-of-COVID-19-cases-in-Ontario."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: " COVID-19 is a respiratory disease first discovered in 2019 and then spread across the world. Many people may be facing an increased level of anxiety and stress. In this report, we built a poisson model to estimate the COVID 19 cases in Ontario. We focus on how gender and age group affect the probability of getting COVID-19. "
output:
  bookdown::pdf_document2
bibliography: references.bib
---

@@

```{r}
unique(clean_data$client_gender)
```

```{r}
sum(clean_data$client_gender=="TRANS WOMAN")
```

```{r}
sum(clean_data$client_gender=="TRANSGENDER")
```

@@

COVID-19 is an infection disease first discovered in Wuhan, China in 2019 and then spread across the world. The disease spreads through small infected respiratory droplets, the infected droplets may direct contact with mouth, nose, or eyes and result in infection. The use of face masks, vaccinations, and social distancing are effective preventative measures for COVID-19. Evidence shows that people who are unvaccinated have a larger death rate than people who are fully vaccinated. [@vaccine] Also, wash your hands often and avoid touching your mouth, nose, and eyes with unwashed hands.

COVID-19 is highly contagious, but the disease has a low fatality rate of less than 6% in the 20 most affected countries. Most infected people will have symptoms like fever, cough, tiredness, etc. These mild symptoms mainly affect the respiratory system and can be recovered in 10 days without hospitalization. If you experience symptoms like shortness of breath, chest pain, etc., you might need to seek emergency medical attention. However, some patients experience lingering health problems even when they have recovered from the acute phase of the illness. Common post-COVID syndrome includes loss or distortion of the sense of smell and taste, fatigue, chest pain, etc.

In this paper, we will focus on factors that increase the probability of getting severe COVID-19. Based on covid-19 cases data obtained from Open Data Toronto, we built a logistic model to predict the probability of COVID-19 patients getting severe symptoms. We found out that aged, male, sporadic patients are more likely to get severe symptoms that need hospitalization. If we keep age and outbreak associated cases fixed, the odds of requiring hospitalization for males are 1.5 times as large as the odds for females. Since we know from this report that male and aged people have a higher risk of severe illness from covid-19, we need to step up our advocacy and raise awareness of the importance of the COVID-19. We urge people to get vaccinated, wear face masks, and minimize going to crowded places.

The article is structured as follows. We first introduced the background knowledge about COVID-19, then we will perform explanatory data analysis. The method section includes all the statistical methods we used in this paper. Then we compare different models and choose a preferred logistic model to estimate the probability of hospitalization of COVID-19 cases. Finally, we will discuss the limitations of our model and future steps.

@@

To attempt an analysis of severe illness from COVID-19, I manipulated COVID-19 cases data. COVID-19 cases in Toronto dataset were managed by Toronto Public Health. These data are retrieved from Open data Toronto. The dataset was gathered in 2020 and it is refreshed and overwritten every Wednesday. It contains demographic and geographic information about every confirmed COVID-19 cases in Toronto.

@@

All analyses were done with a statistic programming language R.[@citeR] We first use R packages "opendatatoronto"[@opendatatoronto] and "tidyverse" [@tidyverse]to import our data. Then I cleaned the names of variables and removed missing values for age. The data was done cleaning.

The raw dataset contains 32000 observations in total. I randomly selected 30000 observations to complete an exploratory analysis and modeling. I also set the seed as 1 so that every time we run the code, we can get the same result. The data was divided into two parts- training and test. Usually, the number of training data is more than the number of detection data, so I randomly selected 70% of data without replacement (21000 observations) as training data. We used the training data for model building and model selection. Then we fitted the optimum logistic model in the test dataset to see whether we can validate the optimum model. Table \@ref(tab:trainsummary) and Table \@ref(tab:testsummary) show a summary of important variables for both the training and test dataset. We can see that the different levels of each variable account for roughly the same proportion of the two datasets, so we believe that there is no significant difference between the training and test dataset. For example, 50.1% of the sample is female in the test dataset, and 49.7% of the sample is female in the training dataset.

@@

The dataset contains demographic (age group, gender) and geographic (patient’s neighborhood name and postal code), and severity (ever hospitalized, ever in ICU, outcome, etc.) information. This information is sensitive and it is possible to identify individuals either directly or indirectly. For example, if you see a case of a transgender patient in your community in the data, combined with age and FSA (first three characters of postal code), it is highly likely that you can identify that person.

First, show an extract of our dataset (figure \@ref(datasummary)). We can tell that most variables are categorical variables.

```{r datasummary,echo=FALSE,warning=FALSE,message=FALSE}
clean_data %>%
  select(reported_date,age_group, client_gender, source_of_infection,neighbourhood_name,ever_hospitalized,outcome)%>%
  slice(1:10) %>%
  kable(
    caption = "A glimpse of important varaibles",
    col.names = c("Reported date","Age group","Gender","Source of infection","Neighbourhood name","Ever hospitalized", "outcome"),
    digits = 1,
    booktabs = TRUE, 
    linesep = "",
    align = c('l', 'l', 'c', 'c', 'c','c', 'r'),
  ) 
```

```{r gender,echo=FALSE,warning=FALSE,message=FALSE}

ggplot(data= clean_data,aes(x = client_gender,fill=ever_hospitalized)) + 
geom_bar(alpha = 0.6,position="identity") +
coord_flip() +
labs(y="count", x="Gender",title = "Gender composition of hospitalized persons") +
theme_minimal() 
```
From figure \@ref(fig:gender), The number of people who receive hospitalization is about the same for women and men.

```{r piegraph,warning=FALSE,message=FALSE,echo = FALSE}
library(plotrix)
hospitcal_data <- clean_data %>%
  filter(ever_hospitalized =="Yes")
names = c("19 and younger",
          "20 to 29 Years", 
          "30 to 39 Years", 
          "40 to 49 Years", 
          "50 to 59 Years", 
          "60 to 69 Years", 
          "70 to 79 Years",
          "80 to 89 Years",
          "90 and older")
# count the number of people for each main source of stress
p1 <- sum(hospitcal_data$age_group =="19 and younger") 
p2 <- sum(hospitcal_data$age_group =="20 to 29 Years")
p3 <- sum(hospitcal_data$age_group =="30 to 39 Years")
p4 <- sum(hospitcal_data$age_group =="40 to 49 Years")
p5 <- sum(hospitcal_data$age_group =="50 to 59 Years")
p6 <- sum(hospitcal_data$age_group =="60 to 69 Years")
p7 <- sum(hospitcal_data$age_group =="70 to 79 Years")
p8 <- sum(hospitcal_data$age_group =="80 to 89 Years")
p9 <- sum(hospitcal_data$age_group =="90 and older")
y=c(p1,p2,p3,p4,p5,p6,p7,p8,p9)
pie3D(y, labels=names,explode=0.1,col = hcl.colors(length(y)),main ="Age composition of severe illness patients",labelcex = 0.75,theta = 1.5)

@@

```{r fatal,warning=FALSE,message=FALSE,echo = FALSE}
clean_data %>%
  filter(ever_hospitalized =="Yes") %>%
  ggplot(aes(x=outcome)) +
  geom_bar(alpha = 0.6,position="identity",col=c("pink","purple"),fill=c("pink","purple")) +
  labs(title="Outcome for COVID-19 patients") +
  theme_minimal()

@@

From figure \@ref(fig:fatal), roughly 40% of all cases that were hospitalized related to their COVID-19 infection are fatal, which indicates there are no effective treatments for patients with severe symptoms of covid.

```{r source,warning=FALSE,message=FALSE,echo = FALSE}
names = c("No Information", 
          "Travel", 
          "Community", 
          "Close Contact", 
          "Pending", 
          "Household Contact", 
          "Outbreaks, Healthcare Institutions",
          "Outbreaks, Other Settings",
          "Outbreaks, Congregate Settings")
# count the number of people for each main source of stress
a1 <- sum(clean_data$source_of_infection =="No Information") 
a2 <- sum(clean_data$source_of_infection =="Travel")
a3 <- sum(clean_data$source_of_infection =="Community")
a4 <- sum(clean_data$source_of_infection =="Close Contact")
a5 <- sum(clean_data$source_of_infection =="Pending")
a6 <- sum(clean_data$source_of_infection =="Household Contact")
a7 <- sum(clean_data$source_of_infection =="Outbreaks, Healthcare Institutions")
a8 <- sum(clean_data$source_of_infection =="Outbreaks, Other Settings")
a9 <- sum(clean_data$source_of_infection =="Outbreaks, Congregate Settings")
a=c(a1,a2,a3,a4,a5,a6,a7,a8)
# calculate the percentage
piepercent = paste(round(100*y/sum(y)), "%")
pie(y, labels=piepercent,col = hcl.colors(length(a), "Spectral"), main ="Source of infection")
legend("topright", names, cex=0.5, fill=hcl.colors(length(a), "Spectral"))
```

From figure \@ref(fig:source), the source of infection in half of the cases is unknown and cannot be speculated. Household contact and close contact with covid positive are also important causes of COVID-19 infection. 10% of COVID patients are infected in healthcare institutions. Healthcare Institutions include hospitals, retirement homes, etc. Usually, people who go to the healthcare institutions for help are mainly the aged population. These people are in poor health conditions and have lower immunity, which makes them more susceptible to viral infections.

@@

To find out what factors can make you seriously ill after getting covid, I performed logistic regression using R. Through the logistic regression model, I was unable to define causality, but rather explore the relationship between different variables. Since patients with mild symptoms can be recovered in a few days without hospitalization, I consider those who are hospitalized as severe illness of COVID-19. I set “ever hospitalized” as the explanatory variable. The variable “ever hospitalized” is a binary variable with two levels: Yes and No, which is an appropriate structure of the outcome variable. I recoded the “ever hospitalized” and set the outcome variable to be 1 if the patient received hospitalization, and 0 if vice versa. The logistic model is constructed using the R package “stats”.[@stats]

The general form of logistic model is :

$log(\frac{\hat{p}}{1-\hat{p}}) = X_i\beta$

@@

@@

$log(\frac{p}{1-p})=\beta_0+\beta_1Client \ gender+\beta_2Age \ group$
The probability of hospitalization is :
$\hat{p}=\frac{e^{\hat{\beta_0}+\hat{\beta_1}Client \ gender+\hat{\beta_2}Age \ group}}{e^{\beta_0+\hat{\beta_1}Client \ gender+\hat{\beta_2}Age \ group}+1}$

### Model 2 (final model)

On the basis of model 1, I want to study the effect of the variable “outbreak associated”. This variable tells us whether the case is associated with a healthcare institution or not.

$log(\frac{p}{1-p}) = \beta_0 + \beta_1Outbreak \ associated + \beta_2Client \ gender + \beta_3Age \ group$

The probability of hospitalization is :
$\hat{p}=\frac{e^{\hat{\beta_0}+ \hat{\beta_1}Outbreak \ associated + \hat{\beta_2}Client \ gender +\hat{\beta_3}Age \ group}}{e^{\hat{\beta_0} + \hat{\beta_1}Outbreak \ associated + \hat{\beta_2}Client \ gender + \hat{\beta_3}Age \ group}+1}$

### Model 3
Based on model 2, I am also interested in whether the interaction term of outbreak associated and gender affect the probability of severe illness of COVID-19.

$log(\frac{p}{1-p}) = \beta_0 + \beta_1Outbreak \ associated + \beta_2Client \ gender + \beta_3Age \ group +\beta_4Client \ gender:Outbreak \ associated$

@@

$\hat{p}=\frac{e^{\hat{\beta_0}+ \hat{\beta_1}Outbreak \ associated + \hat{\beta_2}Client \ gender +\hat{\beta_3}Age \ group+\hat{\beta_4}Client \ gender:Outbreak \ associated}}{e^{\hat{\beta_0} + \hat{\beta_1}Outbreak \ associated + \hat{\beta_2}Client \ gender + \hat{\beta_3}Age \ group+\hat{\beta_4}Client \ gender:Outbreak \ associated}+1}$

After model construction, I will select the best model in the next step. Then model assumptions will be checked and the preferred model is tested with test data to see if it can be validated.

@@

$log(\frac{\hat{p}}{1-\hat{p}}) = -5.5164+0.434Client \ gender_{Male} + 0.5701Age \ group_{20-29 \ years}+0.8237Age \ group_{30-39 \ years}+1.5476Age \ group_{40-49 \ years} +2.0775Age \ group_{50-59 \ years} +3.0685Age \ group_{60-69 \ years} +4.2582Age \ group_{70-79 \ years}+4.6742Age \ group_{80-89 \ years} +4.4333Age \ group_{90 +}$

```{r ,include=FALSE}
mod1 <- glm(hosptial~client_gender + age_group, family = "binomial", data = train)
summary(mod1)
```
Table \@ref(tab:model1) shows a summary of model 1. Most of the predictors are statistically significant at a 0.05 significant level.

@@

$log(\frac{\hat{p}}{1-\hat{p}}) =-5.7373 +0.279 Outbreak \ associated_{Sporadic} + 0.4189	Client \ gender_{Male} + 0.537Age \ group_{20-29 \ years}+0.7959Age \ group_{30-39 \ years}+1.5287Age \ group_{40-49 \ years} +2.0612Age \ group_{50-59 \ years} +3.0585Age \ group_{60-69 \ years} + 4.2752Age \ group_{70-79 \ years}+4.7766Age \ group_{80-89 \ years} +4.61334Age \ group_{90 +}$

Table \@ref(tab:model2) shows a summary of model 2.

@@

$log(\frac{\hat{p}}{1-\hat{p}}) =-5.8792 +0.4807 Outbreak \ associated_{Sporadic} + 0.707	Client \ gender_{Male} + 0.5474Age \ group_{20-29 \ years}+0.8055Age \ group_{30-39 \ years}+1.5366Age \ group_{40-49 \ years} +2.0678Age \ group_{50-59 \ years} +3.0633Age \ group_{60-69 \ years} + 4.2713Age \ group_{70-79 \ years}+4.7898Age \ group_{80-89 \ years} +4.6534Age \ group_{90 +}-0.4029 Outbreak \ associated_{Sporadic}:Client \ gender_{Male}$

```{r ,include=FALSE}
mod3 <- glm(hosptial~outbreak_associated+client_gender+age_group+client_gender:outbreak_associated,  family = "binomial", data = train)
summary(mod3)
```
Table \@ref(tab:model3) shows a summary of model 3.

@@

In this case, model 1 is a subset of model 2, so the two models are nested. Table \@ref(tab:compare12) shows a likelihood ratio test to compare model 1 and model 2. From table \@ref(tab:compare12), the p-value is 0.0012, which is smaller than 0.05, so we have strong evidence against the null hypothesis that the reduced model (mode 1) better explains the COVID-19 cases data. In conclusion, we believe that model 2 (full model) better explains our data.

@@

Model 2 is a subset of model 3, so I performed a likelihood ratio test to compare mode 2 and model 3. (Table \@ref(tab:compare23)) From table \@ref(tab:compare23), the p-value is 0.068, which is greater than 0.05, so we reject the null hypothesis ($H_0$) that model 2 better explains the COVID-19 cases data. Compared to model 3, model 2 better explains the data. In conclusion, model 2 is the optimum model.

@@

# Model interpretation
The optimum model is :

$log(\frac{\hat{p}}{1-\hat{p}}) =-5.7373 +0.279 Outbreak \ associated_{Sporadic} + 0.4189	Client \ gender_{Male} + 0.537Age \ group_{20-29 \ years}+0.7959Age \ group_{30-39 \ years}+1.5287Age \ group_{40-49 \ years} +2.0612Age \ group_{50-59 \ years} +3.0585Age \ group_{60-69 \ years} + 4.2752Age \ group_{70-79 \ years}+4.7766Age \ group_{80-89 \ years} +4.61334Age \ group_{90 +}$

>>>>

---
title: "Elderly, Male, Sporadic Population is More Likely to Get Severe COVID-19"
subtitle: "A estimation of the probability of developing severe COVID-19 using logistic model"
author:
- Zihan Zhang
thanks: "Code and data are available at: https://github.com/zihanjasmine/Estimation-of-the-probability-of-severe-illness-of-COVID-19"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "COVID-19 is a respiratory disease first discovered in 2019 and then spread across the world. In this report, a logistic model was built using the programming language R to estimate the probability of severe illness of COVID. The results show that elderly, male population is at high risk for developing severe COVID-19 symptoms. This article illustrates the high-risk population for COVID-19 and suggests a new direction for studying the causes of severe COVID-19 that requires hospitalization.
 \\par
 \\textbf {Keywords:}  logistic regression, severe illness of COVID-19, open data toronto" 
output:
  bookdown::pdf_document2
bibliography: references.bib
---

@@

@@

COVID-19 is an infection disease first discovered in Wuhan, China, in 2019, which then spread across the world. The disease primarily spreads through the direct contact of small infected respiratory droplets with open mucus membranes, such as mouth, nose, or eyes. The use of face masks, vaccinations, and social distancing are effective preventative measures for COVID-19. Evidence shows that people who are unvaccinated have a higher death rate than people who are fully vaccinated [@vaccine]. Moreover, washing your hands often and avoid touching your mouth, nose, and eyes also reduces the likelihood of being infected.

COVID-19 is highly contagious, but the disease has a low fatality rate of less than 6% in the 20 most affected countries. Most infected people will develop symptoms such as fever, cough, tiredness, etc. These mild symptoms mainly affect the respiratory system, and those affected can recover in 10 days without hospitalization. If you experience symptoms like shortness of breath, chest pain, etc., you might need to seek emergency medical attention. However, some patients experience lingering health problems even when they have recovered from the acute phase of the illness. Some common post-COVID syndromes include loss or distortion of the sense of smell and taste, fatigue, chest pain, etc.

In this paper, I will focus on factors that increase the probability of developing severe COVID-19 symptoms. Based on COVID-19 cases data obtained from Open Data Toronto, a logistic model was built to predict the probability of COVID-19 patients getting severe symptoms. The evidence suggests that aged, male, and sporadic patients are more likely to develop severe symptoms that require hospitalization. Sporadic patients means the patients' virus infection is not associated with healthcare institutions. If age and outbreak associated cases remains the same, the odds of males requiring hospitalization are 1.5 times larger than the odds for females. Since this report suggests that male and aged people have a higher risk of severe illness from COVID-19, more efforts should be made to step up advocacy and raise awareness of the importance of COVID-19 preventative measures. I urge people to get vaccinated, wear face masks, and minimize going to crowded places.

The article is structured as follows. This paper first provides an introduction to the background knowledge about COVID-19. Then data collection, data processing and data charactistic are included in data section. The model section is consists of model assumption, model construction and model validation. Model selection and interpretation are shown in the result section. Finally, discussion is comprised of three parts: summary, findings, and limitation & future work.

@@

In attempts to generate an analysis of severe symptom of COVID-19, I manipulated "COVID-19 Cases in Toronto" dataset, managed by Toronto Public Health. The data is retrieved from Open data Toronto. The dataset was gathered in 2020 and it is being refreshed and overwritten every Wednesday. It contains demographic and geographic information about every confirmed COVID-19 cases in Toronto.

@@

All analyses were done with a statistic programming language R [@citeR]. I first used R packages "opendatatoronto"[@opendatatoronto] and "tidyverse" [@tidyverse]to to import data. Then I cleaned the names of variables and removed missing values for age. The data was done cleaning. R packages "ggplot2" [@ggplot2] and "KableExtra" [@kableextra] were used for data visualization.

The raw dataset contains 32000 observations in total. I randomly selected 30000 observations to complete an exploratory analysis and modeling. I also set the seed as 1 so that the same result can be obtained every time I run the code.

The data was divided into two parts- training and test. Usually, the number of observations in the training dataset is larger than the number of observations in the test dataset, so I randomly selected 70% of data without replacement (21000 observations) as training data. The training data was used for model building and model selection. Then I fitted the optimum logistic model in the test dataset to see whether the optimum model can be validated. Table \@ref(tab:trainsummary) and Table \@ref(tab:testsummary) show a summary of important variables for both the training and test dataset. The different levels of each variable account for roughly the same proportion of the two datasets, so I believe that there is no significant difference between the training and test dataset. For example, 50.1% of the sample is female in the test dataset, and 49.7% of the sample is female in the training dataset.

@@

The dataset contains demographic (age group, gender) and geographic (patient’s neighborhood name and postal code), and severity (ever hospitalized, ever in ICU, outcome, etc.) information. This information is sensitive, and it is possible to identify individuals either directly or indirectly. For example, if you see a case of a transgender patient in your community in the data, combined with age and FSA (first three characters of postal code), it is highly likely that you can identify that person.

```{r datasummary,echo=FALSE,warning=FALSE,message=FALSE,fig.cap="A glimpse of important varaibles",fig.height = 3, fig.width = 5}
clean_data %>%
  select(reported_date,age_group, source_of_infection,neighbourhood_name,ever_hospitalized,outcome)%>%
  slice(1:8) %>%
  kable(
    caption = "A glimpse of important varaibles",
    col.names = c("Reported date","Age group","Source of infection","Neighbourhood name","Ever hospitalized", "outcome"),
    digits = 1,
    booktabs = TRUE, 
    linesep = "",
    align = c('l', 'l', 'c', 'c','c', 'r'),
  ) 
```

Table \@ref(tab:datasummary) tells that most variables are categorical variables as suggested by an extract of our dataset.

```{r gender,echo=FALSE,warning=FALSE,message=FALSE,fig.cap="Gender composition of hospitalized persons",fig.height = 3, fig.width = 5}
ggplot(data= clean_data,aes(x = client_gender,fill=ever_hospitalized)) + 
geom_bar(alpha = 0.6,position="identity") +
coord_flip() +
labs(y="count", x="Gender",title = "Gender composition of hospitalized persons") +
theme_minimal() 
```

Figure \@ref(fig:gender) indicates that the number of people receiving hospitalization is about the same for women and men. After careful consideration, gender is kept as a binary variable for males and females only for modeling. There are only three observations for transgender patients, which might lead to high instability in our model.

```{r piegraph,warning=FALSE,message=FALSE,echo = FALSE,fig.cap="Age composition of severe illness patients",fig.height = 4, fig.width = 5.5}
library(plotrix)
hospitcal_data <- clean_data %>%
  filter(ever_hospitalized =="Yes")
names = c("19 and younger",
          "20 to 29 Years", 
          "30 to 39 Years", 
          "40 to 49 Years", 
          "50 to 59 Years", 
          "60 to 69 Years", 
          "70 to 79 Years",
          "80 to 89 Years",
          "90 and older")
# count the number of people for each main source of stress
p1 <- sum(hospitcal_data$age_group =="19 and younger") 
p2 <- sum(hospitcal_data$age_group =="20 to 29 Years")
p3 <- sum(hospitcal_data$age_group =="30 to 39 Years")
p4 <- sum(hospitcal_data$age_group =="40 to 49 Years")
p5 <- sum(hospitcal_data$age_group =="50 to 59 Years")
p6 <- sum(hospitcal_data$age_group =="60 to 69 Years")
p7 <- sum(hospitcal_data$age_group =="70 to 79 Years")
p8 <- sum(hospitcal_data$age_group =="80 to 89 Years")
p9 <- sum(hospitcal_data$age_group =="90 and older")
y=c(p1,p2,p3,p4,p5,p6,p7,p8,p9)
pie3D(y, labels=names,explode=0.1,col = hcl.colors(length(y)),main ="Age composition of severe illness patients",labelcex = 0.6,theta = 1.5)

@@

```{r fatal,warning=FALSE,message=FALSE,echo = FALSE,fig.cap="Outcome for COVID-19 patients",fig.height = 2.5, fig.width = 3.5}
clean_data %>%
  filter(ever_hospitalized =="Yes") %>%
  ggplot(aes(x=outcome)) +
  geom_bar(alpha = 0.6,position="identity",col=c("pink","purple"),fill=c("pink","purple")) +
  labs(title="Outcome for COVID-19 patients") +
  theme_minimal()

@@

From figure \@ref(fig:fatal), roughly 40% cases that were hospitalized are fatal, which indicates there are no effective treatments for patients with severe symptoms of COVID-19.

```{r source,warning=FALSE,message=FALSE,echo = FALSE,fig.cap="Source of infection",fig.height = 5, fig.width = 6.5}
names = c("No Information", 
          "Travel", 
          "Community", 
          "Close Contact", 
          "Pending", 
          "Household Contact", 
          "Outbreaks, Healthcare Institutions",
          "Outbreaks, Other Settings",
          "Outbreaks, Congregate Settings")
# count the number of people for each main source of stress
a1 <- sum(clean_data$source_of_infection =="No Information") 
a2 <- sum(clean_data$source_of_infection =="Travel")
a3 <- sum(clean_data$source_of_infection =="Community")
a4 <- sum(clean_data$source_of_infection =="Close Contact")
a5 <- sum(clean_data$source_of_infection =="Pending")
a6 <- sum(clean_data$source_of_infection =="Household Contact")
a7 <- sum(clean_data$source_of_infection =="Outbreaks, Healthcare Institutions")
a8 <- sum(clean_data$source_of_infection =="Outbreaks, Other Settings")
a9 <- sum(clean_data$source_of_infection =="Outbreaks, Congregate Settings")
a=c(a1,a2,a3,a4,a5,a6,a7,a8)
# calculate the percentage
piepercent = paste(round(100*y/sum(y)), "%")
pie(y, labels=piepercent,col = hcl.colors(length(a), "Spectral"), main ="Source of infection")
legend("topright", names, cex=0.5, fill=hcl.colors(length(a), "Spectral"))
```

From figure \@ref(fig:source), the source of infection in half of the cases is unknown and cannot be speculated. Household contact and close contact with covid positive are also important causes of COVID-19 infection. 10% of COVID patients are infected in healthcare institutions. Healthcare Institutions include hospitals, retirement homes, etc. Usually, people who go to the healthcare institutions for help are mainly the aged population. These people are in poor health conditions and have a weaker immunity, making them more susceptible to viral infections.

@@

To find out what factors can make you seriously ill after getting covid, I performed logistic regression using R [@citeR]. Through the logistic regression model, I am unable to define causality, but rather explore the relationship between different variables. Since patients with mild symptoms can be recovered in a few days without hospitalization, I consider those who are hospitalized as severe illness of COVID-19. I set “ever hospitalized” as the outcome variable. The variable “ever hospitalized” is a binary variable with two levels: Yes and No, which is an appropriate structure of the outcome variable. I recoded the “ever hospitalized” and set the outcome variable to be 1 if the patient received hospitalization, and 0 if vice versa. The logistic model is constructed using the R package “stats” [@stats].

The general form of logistic model is (see Equation \@ref(eq:log)) :

\begin{equation} 
  log(\frac{\hat{p}}{1-\hat{p}}) = X_i\beta
  (\#eq:log)
\end{equation}

@@

## model assumption
Before model construction, I will check whether the assumptions of the model are satisfied. It is important to check the assumptions of the model because if the assumptions of the model are not satisfied, it will lead to inaccurate predictions.

There are four assumptions for a logistic model.

- Assumption of appropriate outcome type

Logistic regression assumes the outcome variable to be binary with two levels. The response variable should follow binomial or bernoulli distribution.

- Assumption of independence of observations

Logistic regression assumes that the observations must be independent of each other. Each observation should represent one patient, so no repeated data is measured. In this paper, the residual plot will be used to check the independence assumption. If the residuals in the plot are randomly scattered around 0, the independence assumption is satisfied.

- Assumption of absence of multicollinearity

Multicollinearity refers to a strong linear relationship between predictors, which might result in non-significant predictors, weakened statistical power of our logistic model, etc. Multicollinearity is accessed through VIF value, a predictor with a VIF value that exceeds 5 indicates strong multicollinearity, which suggests the predictor might need to be removed.

- Assumption of large sample size

Logistic model assumes there are sufficient observations for each variable to avoid overfitting.

@@

The first model is (see Equation \@ref(eq:1)):

\begin{equation} 
  log(\frac{p}{1-p})=\beta_0+\beta_1Client \ gender+\beta_2Age \ group
  (\#eq:1)
\end{equation}

The probability of hospitalization is (see Equation \@ref(eq:1a)):

\begin{equation} 
  \hat{p}=\frac{e^{\hat{\beta_0}+\hat{\beta_1}Client \ gender+\hat{\beta_2}Age \ group}}{e^{\beta_0+\hat{\beta_1}Client \ gender+\hat{\beta_2}Age \ group}+1}
  (\#eq:1a)
\end{equation}

### Model 2

On the basis of model 1, I want to study the effect of the variable “outbreak associated”. This variable tells us whether the case is associated with healthcare institutions or not. If the infection is related to healthcare institutions, it is outbreak associated and if the infection is not hospital related, it is sporadic.

The second model is (see Equation \@ref(eq:2)) :

\begin{equation} 
  log(\frac{p}{1-p}) = \beta_0 + \beta_1Outbreak \ associated + \beta_2Client \ gender + \beta_3Age \ group
  (\#eq:2)
\end{equation}

The probability of hospitalization is (see Equation \@ref(eq:2a)):

\begin{equation} 
  \hat{p}=\frac{e^{\hat{\beta_0}+ \hat{\beta_1}Outbreak \ associated + \hat{\beta_2}Client \ gender +\hat{\beta_3}Age \ group}}{e^{\hat{\beta_0} + \hat{\beta_1}Outbreak \ associated + \hat{\beta_2}Client \ gender + \hat{\beta_3}Age \ group}+1}
  (\#eq:2a)
\end{equation}

### Model 3
Based on model 2, I am also interested in whether the interaction term of outbreak associated and gender affect the probability of developing severe COVID-19 that requires hospitalization.

The third model is (see Equation \@ref(eq:3)) :

\begin{equation} 
  log(\frac{p}{1-p}) = \beta_0 + \beta_1Outbreak \ associated + \beta_2Client \ gender + \beta_3Age \ group +\beta_4Client \ gender:Outbreak \ associated
  (\#eq:3)
\end{equation}

@@

\begin{equation} 
  \hat{p}=\frac{e^{\hat{\beta_0}+ \hat{\beta_1}Outbreak \ associated + \hat{\beta_2}Client \ gender +\hat{\beta_3}Age \ group+\hat{\beta_4}Client \ gender:Outbreak \ associated}}{e^{\hat{\beta_0} + \hat{\beta_1}Outbreak \ associated + \hat{\beta_2}Client \ gender + \hat{\beta_3}Age \ group+\hat{\beta_4}Client \ gender:Outbreak \ associated}+1}
  (\#eq:3a)
\end{equation}

After model construction, I will select the best model in the next step. The preferred model will be tested with test data to see if it can be validated.

## Model validation

Data validation is an important process to be carried out after model selection, it is a way to evaluate the effectiveness of our optimum model. The COVID-19 cases in Toronto dataset was separated into training and test dataset in the beginning, the training dataset is used for model building and model selection, and the test dataset is used for model validation.

The test dataset will be used to fit the optimum model. If the following conditions are satisfied, it is safe to say the optimum model is validated. First, the difference of model coefficients of the test dataset is within the standard error of corresponding coefficients of the training dataset. Second, the predictors of the optimum model appear significant in both training and test datasets. Thirdly, when fitting the optimum model using test dataset, the test dataset must not have more serious assumption violations.

@@

\begin{equation} 
\begin{split}
log(\frac{\hat{p}}{1-\hat{p}}) = -5.5164+0.434Client \ gender_{Male} + 0.5701Age \ group_{20-29 \ years}+0.8237Age \ group_{30-39 \ years}\\
+1.5476Age \ group_{40-49 \ years} +2.0775Age \ group_{50-59 \ years}+3.0685Age \ group_{60-69 \ years}\\
+4.2582Age \ group_{70-79 \ years}+4.6742Age \ group_{80-89 \ years} +4.4333Age \ group_{90 +}\\
\end{split}
(\#eq:mod1)
\end{equation}

```{r ,include=FALSE}
mod1 <- glm(hosptial~client_gender + age_group, family = "binomial", data = train)
summary(mod1)
```
Table \@ref(tab:model1) shows a summary of model 1 (see Equation \@ref(eq:mod1)). Most of the predictors are statistically significant at a 0.05 significant level.

@@

\begin{equation} 
\begin{split}
log(\frac{\hat{p}}{1-\hat{p}}) = -5.7373 +0.279 Outbreak \ associated_{Sporadic} + 0.4189	Client \ gender_{Male} + 0.537Age \ group_{20-29 \ years}\\
+0.7959Age \ group_{30-39 \ years}+1.5287Age \ group_{40-49 \ years} +2.0612Age \ group_{50-59 \ years}+3.0585Age \ group_{60-69 \ years}\\
+ 4.2752Age \ group_{70-79 \ years}+4.7766Age \ group_{80-89 \ years} +4.61334Age \ group_{90 +}\\
\end{split}
(\#eq:mod2)
\end{equation}

Table \@ref(tab:model2) shows a summary of model 2 (see Equation \@ref(eq:mod2)).

@@

\begin{equation} 
\begin{split}
log(\frac{\hat{p}}{1-\hat{p}}) = -5.8792 +0.4807 Outbreak \ associated_{Sporadic} + 0.707	Client \ gender_{Male} + 0.5474Age \ group_{20-29 \ years}\\
+0.8055Age \ group_{30-39 \ years}+1.5366Age \ group_{40-49 \ years} +2.0678Age \ group_{50-59 \ years}+3.0633Age \ group_{60-69 \ years}\\
+ 4.2713Age \ group_{70-79 \ years}+4.7898Age \ group_{80-89 \ years} +4.6534Age \ group_{90 +}\\
-0.4029 Outbreak \ associated_{Sporadic}:Client \ gender_{Male}\\
\end{split}
(\#eq:mod3)
\end{equation}

```{r ,include=FALSE}
mod3 <- glm(hosptial~outbreak_associated+client_gender+age_group+client_gender:outbreak_associated,  family = "binomial", data = train)
summary(mod3)
```
Table \@ref(tab:model3) shows a summary of model 3 (see Equation \@ref(eq:mod3)).

@@

In this case, model 1 is a subset of model 2, so the two models are nested. Table \@ref(tab:compare12) shows a likelihood ratio test to compare model 1 and model 2. From table \@ref(tab:compare12), the p-value is 0.0012, which is smaller than 0.05, so I have strong evidence against the null hypothesis that the reduced model (mode 1) better explains the COVID-19 cases data. In conclusion,compared to model 1, model 2 (full model) better explains the data.

@@

Model 2 is a subset of model 3, so I performed a likelihood ratio test to compare mode 2 and model 3. (Table \@ref(tab:compare23)) From table \@ref(tab:compare23), the p-value is 0.068, which is greater than 0.05, so I failed to reject the null hypothesis ($H_0$) that model 2 better explains the COVID-19 cases data. Compared to model 3, model 2 better explains the data. In conclusion, model 2 (see Equation \@ref(eq:mod2)) is the optimum model.

@@

## Model interpretation
The optimum model is :

\begin{equation} 
\begin{split}
log(\frac{\hat{p}}{1-\hat{p}}) = -5.7373 +0.279 Outbreak \ associated_{Sporadic} + 0.4189	Client \ gender_{Male} + 0.537Age \ group_{20-29 \ years}\\
+0.7959Age \ group_{30-39 \ years}+1.5287Age \ group_{40-49 \ years} +2.0612Age \ group_{50-59 \ years}+3.0585Age \ group_{60-69 \ years}\\
+ 4.2752Age \ group_{70-79 \ years}+4.7766Age \ group_{80-89 \ years} +4.61334Age \ group_{90 +}\\
\end{split}
(\#eq:finalmod)
\end{equation}

- Outbreak associated

When all other predictors except outbreak associated remain the same, the odds of hospitalization for an outbreak-associated patient is 1.3218 ($e^{0.279}$) times larger than the odds of hospitalization for a sporadic patient.

- Gender

When all other predictors except gender remain the same, the odds of hospitalization for a male is 1.5203 times larger than the odds for a female being admitted to a hospital.

- Age group
